Galaxy Zoo: the dependence of the star formation-stellar mass relation on spiral disk morphology

### 2015-02-09 paper/mainsequence_morph.tex
Adding color for ref response highlights
Highlighted changes in text for referee response

\newcommand{\mytilde}{\textasciitilde}
\newcommand{\kevin}{{\color{red}\textbf{Kevin}}}

@@

\begin{abstract}
We measure the stellar mass-star formation rate relation in star-forming disk galaxies at $z\le0.085$, using Galaxy~Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their relative pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is significantly smaller than the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the various quenching mechanisms or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed, but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

Stellar masses and star formation rates are computed from optical diagnostics and taken from the MPA-JHU catalogue \citep{kau03a,bri04,sal07}. We use updated masses and activity classifications from the DR7 database.\footnote{\url{http://home.strw.leidenuniv.nl/\mytilde jarle/SDSS/}} We select only galaxies with $M_\star > 10^8 M_\odot$, for which \citet{bri04} estimate that the MPA-JHU sample is complete. Star-forming galaxies are separated from other emission-line galaxies using the standard BPT classification \citep*{bal81} below the \citet{kau03} demarcation. Galaxies classified as star-forming but with low signal-to-noise ratio $(S/N > 3)$ are also excluded. Both $M_\star$ and SFR are measured using median values extracted from the probability distribution functions.

The spiral nature of the star-forming disk galaxies is identified according to the following thresholds in the GZ2 spectroscopic sample, where $p$ is the debiased vote fraction and $N$ the weighted number of total votes: $p_\textrm{features/disk} > 0.430$, $p_\textrm{not~edgeon} > 0.715$, $p_\textrm{spiral}>0.619$, and $N_\textrm{spiral}>20$. These cuts are chosen to ensure reliable identification and with enough data points such that spiral substructure has a reasonable estimate of the associated uncertainty. Sub-classes of spiral structure are identified by weighting each galaxy according to the fraction of votes received in each morphological category.

\begin{table} 
 \begin{tabular}{@{}lrcrcl}
 \hline
\multicolumn{1}{c}{Sample} &
\multicolumn{1}{c}{$N$} &
\multicolumn{1}{c}{$\alpha$} &
\multicolumn{1}{c}{$\beta$} &
\multicolumn{1}{c}{$\sigma_\alpha$} &
\multicolumn{1}{c}{$\sigma_\beta$} 
\\ 
\hline
\hline						
SF galaxies  & 48405  & $0.72$  & $-7.08$  &  $7.13\times10^{-6}$  & $6.98\times10^{-4}$  \\
\hline
1 arm        & 289    & $0.74$  & $-7.18$  &  $2.70\times10^{-5}$  & $2.69\times10^{-3}$  \\
2 arms       & 5641   & $0.78$  & $-7.69$  &  $3.70\times10^{-5}$  & $3.78\times10^{-3}$  \\
3 arms       & 997    & $0.71$  & $-6.99$  &  $4.49\times10^{-5}$  & $4.65\times10^{-3}$  \\
4 arms       & 285    & $0.71$  & $-7.05$  &  $4.80\times10^{-5}$  & $4.93\times10^{-3}$  \\
5+ arms      & 289    & $0.80$  & $-8.11$  &  $7.54\times10^{-5}$  & $7.53\times10^{-3}$  \\
can't tell   & 2017   & $0.77$  & $-7.72$  &  $4.53\times10^{-5}$  & $4.54\times10^{-3}$  \\
\hline
tight arms   & 3255   & $0.70$  & $-6.99$  &  $1.79\times10^{-3}$  & $1.86\times10^{-1}$  \\
medium arms  & 4578   & $0.77$  & $-7.62$  &  $6.81\times10^{-3}$  & $6.87\times10^{-1}$  \\
loose arms   & 1672   & $0.73$  & $-7.08$  &  $5.81\times10^{-3}$  & $5.76\times10^{-2}$  \\
\hline
barred       & 3199   & $0.75$  & $-7.38$  &  $8.53\times10^{-5}$  & $8.42\times10^{-3}$  \\
unbarred     & 11846  & $0.69$  & $-6.78$  &  $2.91\times10^{-5}$  & $2.89\times10^{-3}$  \\
\hline
merger       & 2978   & --      & $-6.73$  &  --                   & --                   \\
\hline
 \end{tabular}
\caption{Basic properties of the $M_\star-SFR$ linear fit for GZ2 star-forming galaxies. $N$ is the number of galaxies with plurality classifications for spiral arm multiplicity and pitch angles, and at a cutoff of $p_{bar}=0.4$ for barred/unbarred galaxies. $\alpha$ and $\beta$ are fit according to Equation~\ref{eqn-linearfit} to data weighted by morphological vote fractions for spiral arm multiplicity and pitch angle, and to subsamples split by morphology for barred/unbarred and merging galaxies.\label{tbl-fits}}
\end{table}

@@

\noindent where $\alpha$ and $\beta$ represent the slope and offset, respectively. The formal uncertainties $\sigma_\alpha$ and $\sigma_\beta$ are taken from the covariance matrix for each least-squares fit (Table~\ref{tbl-fits}). In fitting subsamples selected by morphology, we apply the same fit to all star-forming galaxies, but weighted by the morphological likelihood in the GZ2 data. The low number of high-mass galaxies in this volume also means that we are insensitive to possible turnovers in the SFMS at $M_\star>10^{10}~M_\odot$ \citep{whi14,lee15}, emphasizing our choice to fit a linear model.

@@

The sample of galaxies examined here is not explicitly constructed to be volume-limited (although \citealt{bri04} estimate that SFR and $M_\star$ are essentially complete for the mass and $S/N$ limits employed). The main reason for this is that we are comparing effects between sub-samples of galaxies using the same selection functions. One possibility is that volume-limiting would deal with galaxies in which the dust content is high enough to obscure all emission lines, even in the presence of significant star formation. Such galaxies, typically [U]LIRGs, have very low space-densities at $z<0.1$, and typically lack the regular disk structure needed to categorize it for the morphologies considered here.

@@

As a control sample, we also analyze the fits to the underlying star-forming main sequence relation for star-forming disks as measured in a sample of local SDSS galaxies (Figures~\ref{fig-number}-\ref{fig-bar}, \ref{fig-mergers}). As previously demonstrated with SDSS data \citep[e.g.,][]{bri04}, there is a tight correlation between $M_\star$ and SFR, with galaxies in the process of quenching lowering their SFR and falling below the trend. The relationship extends over at least 3~orders of magnitude in both $M_\star$ and SFR. Fits to the SFMS for all subsamples in this paper are listed in Table~\ref{tbl-fits}.

@@

Figure~\ref{fig-number} shows the SFR as a function of $M_\star$ for disk galaxies separated by their arm multiplicity. The GZ2 data separates disk galaxies with visible spiral arms into categories of 1, 2, 3, 4, or more than four spiral arms; there is also an additional option if the number of spiral arms cannot be accurately determined. For this analysis, galaxies in each mass/SFR bin is weighted by the vote fraction for the morphology being tested. The fit to the star-forming main sequence for all six sub-samples of spiral galaxies split by arm multiplicity tightly follows that of the total star-forming disk population. Both the slopes and offsets of each linear fit are consistent within the formal fitting errors in Table~\ref{tbl-fits}. %There are too few individual galaxies with confirmed spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all arm multiplicities at $10^8<M_\star/M_\odot<10^{11}$.

@@

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=10,035$) are dominated by two-armed spirals, which are the only spiral multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\odot$ are detected. The results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compare the sample standard deviation ($\sigma_{SFR}$) to the star-forming galaxy population over its various morphological subsamples. The value of $\sigma_{SFR}$ monotonically decreases with increasing mass over the range $8.0<\log(M/M_\odot)<11.5$. For all morphological populations examined in this paper, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

@@

We have also examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the morphologically-selected categories as compared to the general SFMS.

@@

\citet{abr14} found that by normalizing galaxies by the stellar mass of the disk alone, the slope of the SFMS is consistent with only a linear trend (removing any dependence on mass). Although this correction to the disk stellar mass homogenizes the SFMS for disks with a range of $B/T$, the intrinsic dispersion ($\sigma_{SFR}$) of the sequence must be a result of contributions by bars, disk dynamics, halo heating, AGN activity, environment and/or gas accretion history, among other factors \citep{dut10}. Our results show that the neither of the first two factors play dominant roles in controlling $\sigma_{SFR}$, at least as far as major dynamical drivers (such as strong bars or additional arms) are concerned. Thus while the overall bulge strength does affect the position of a galaxy on the SFMS \citep{mar09,che12a,fan13,kav14,lan14,oma14}, the structure of the \emph{disk itself} does not. This is also consistent with recent models in which details of the feedback, which also relate strongly to the galaxy properties, have little effect on the SFMS \citep{hop14}. Alternatively, this also agrees with models in which the SFMS is the result of stochastic processes, rather than deterministic physics related to galaxy evolution \citep{kel14b}.

@@

We analyze for the first time the detailed structure of disks in large samples of star-forming galaxies in the local Universe as related to their position on the $M_\star$-SFR relation. This analysis is made possible by using morphological classifications from the Galaxy~Zoo~2 project. We find that neither the slope nor the dispersion of the star-forming galaxies are affected when splitting the sample into different categories of disks, including barred/unbarred galaxies, the pitch angle of spiral arms, or the number of spiral arms.

>>>>

\newcommand{\mytilde}{\textasciitilde}

% Highlight referee response
\definecolor{titlecol}{rgb}{0,0,1}
\def\rr{\color{titlecol}}

@@

\begin{abstract}
We measure the {\rr stellar mass-star formation rate relation in star-forming disk galaxies at $z\le0.085$, using Galaxy~Zoo morphologies to examine different populations of spirals as classified by their kiloparsec-scale structure. We examine the number of spiral arms, their relative} pitch angle, and the presence of a galactic bar in the disk, and show that both the slope and dispersion of the $M_\star$-SFR relation is constant when varying all the above parameters. We also show that mergers (both major and minor), which represent the strongest conditions for increases in star formation at a constant mass, only boost the SFR above the main relation by $\sim0.3$~dex; this is significantly smaller than the increase seen in merging systems at $z>1$. Of the galaxies lying significantly above the $M_\star$-SFR relation in the local Universe, more than $50\%$ are mergers. We interpret this as evidence that the spiral arms, which are imperfect reflections of the galaxy's current gravitational potential, are either fully independent of the various quenching mechanisms or are completely overwhelmed by the combination of outflows and feedback. The arrangement of the star formation can be changed, but the system \emph{as a whole} regulates itself even in the presence of strong dynamical forcing. 
\end{abstract}

@@

Stellar masses and star formation rates are computed from optical diagnostics and taken from the MPA-JHU catalogue \citep{kau03a,bri04,sal07}. We use updated masses and activity classifications from the DR7 database.\footnote{\url{http://home.strw.leidenuniv.nl/\mytilde jarle/SDSS/}} {\rr We select only galaxies with $M_\star > 10^8 M_\odot$, for which \citet{bri04} estimate that the MPA-JHU sample is complete. Star-forming galaxies are separated from other emission-line galaxies using the standard BPT classification \citep*{bal81} below the \citet{kau03} demarcation. Galaxies classified as star-forming but with low signal-to-noise ratio $(S/N > 3)$ are also excluded. Both $M_\star$ and SFR are measured using median values extracted from the probability distribution functions.

The spiral nature of the star-forming disk galaxies is identified according to the following thresholds in the GZ2 spectroscopic sample, where $p$ is the debiased vote fraction and $N$ the weighted number of total votes: $p_\textrm{features/disk} > 0.430$, $p_\textrm{not~edgeon} > 0.715$, $p_\textrm{spiral}>0.619$, and $N_\textrm{spiral}>20$. These cuts are chosen to ensure reliable identification and with enough data points such that spiral substructure has a reasonable estimate of the associated uncertainty. Sub-classes of spiral structure are identified by weighting each galaxy according to the fraction of votes received in each morphological category. }

\begin{table} 
 \begin{tabular}{@{}lrcrcl}
 \hline
\multicolumn{1}{c}{Sample} &
\multicolumn{1}{c}{$N$} &
\multicolumn{1}{c}{$\alpha$} &
\multicolumn{1}{c}{$\beta$} &
\multicolumn{1}{c}{$\sigma_\alpha$} &
\multicolumn{1}{c}{$\sigma_\beta$} 
\\ 
\hline
\hline						
SF galaxies  & 48405  & $0.72$  & $-7.08$  &  $7.13\times10^{-6}$  & $6.98\times10^{-4}$  \\
\hline
1 arm        & 289    & $0.74$  & $-7.18$  &  $2.70\times10^{-5}$  & $2.69\times10^{-3}$  \\
2 arms       & 5641   & $0.78$  & $-7.69$  &  $3.70\times10^{-5}$  & $3.78\times10^{-3}$  \\
3 arms       & 997    & $0.71$  & $-6.99$  &  $4.49\times10^{-5}$  & $4.65\times10^{-3}$  \\
4 arms       & 285    & $0.71$  & $-7.05$  &  $4.80\times10^{-5}$  & $4.93\times10^{-3}$  \\
5+ arms      & 289    & $0.80$  & $-8.11$  &  $7.54\times10^{-5}$  & $7.53\times10^{-3}$  \\
can't tell   & 2017   & $0.77$  & $-7.72$  &  $4.53\times10^{-5}$  & $4.54\times10^{-3}$  \\
\hline
tight arms   & 3255   & $0.70$  & $-6.99$  &  $1.79\times10^{-3}$  & $1.86\times10^{-1}$  \\
medium arms  & 4578   & $0.77$  & $-7.62$  &  $6.81\times10^{-3}$  & $6.87\times10^{-1}$  \\
loose arms   & 1672   & $0.73$  & $-7.08$  &  $5.81\times10^{-3}$  & $5.76\times10^{-2}$  \\
\hline
barred       & 3199   & $0.75$  & $-7.38$  &  $8.53\times10^{-5}$  & $8.42\times10^{-3}$  \\
unbarred     & 11846  & $0.69$  & $-6.78$  &  $2.91\times10^{-5}$  & $2.89\times10^{-3}$  \\
\hline
merger       & 2978   & --      & $-6.73$  &  --                   & --                   \\
\hline
 \end{tabular}
\caption{Basic properties of the $M_\star-SFR$ linear fit for GZ2 star-forming galaxies. $N$ is the number of galaxies with plurality classifications for spiral arm multiplicity and pitch angles, and at a cutoff of $p_{bar}=0.4$ for barred/unbarred galaxies. {\rr $\alpha$ and $\beta$ are fit according to Equation~\ref{eqn-linearfit} to data weighted by morphological vote fractions for spiral arm multiplicity and pitch angle, and to subsamples split by morphology for barred/unbarred and merging galaxies.}\label{tbl-fits}}
\end{table}

@@

\noindent where $\alpha$ and $\beta$ represent the slope and offset, respectively. {\rr The formal uncertainties $\sigma_\alpha$ and $\sigma_\beta$ are taken from the covariance matrix for each least-squares fit (Table~\ref{tbl-fits}). In fitting subsamples selected by morphology, we apply the same fit to all star-forming galaxies, but weighted by the morphological likelihood in the GZ2 data. The low number of high-mass galaxies in this volume also means that we are insensitive to possible turnovers in the SFMS at $M_\star>10^{10}~M_\odot$ \citep{whi14,lee15}, emphasizing our choice to fit a linear model.

@@

The sample of galaxies examined here is not explicitly constructed to be volume-limited (although \citealt{bri04} estimate that SFR and $M_\star$ are essentially complete for the mass and $S/N$ limits employed). The main reason for this is that we are comparing effects between sub-samples of galaxies using the same selection functions. One possibility is that volume-limiting would deal with galaxies in which the dust content is high enough to obscure all emission lines, even in the presence of significant star formation. Such galaxies, typically [U]LIRGs, have very low space-densities at $z<0.1$, and typically lack the regular disk structure needed to categorize it for the morphologies considered here.}

@@

As a control sample, we also analyze the fits to the underlying star-forming main sequence relation for star-forming disks as measured in a sample of local SDSS galaxies (Figures~\ref{fig-number}-\ref{fig-bar}, \ref{fig-mergers}). As previously demonstrated with SDSS data \citep[e.g.,][]{bri04}, there is a tight correlation between $M_\star$ and SFR, with galaxies in the process of quenching lowering their SFR and falling below the trend. The relationship extends over at least {\rr 3~orders of magnitude} in both $M_\star$ and SFR. Fits to the SFMS for all subsamples in this paper are listed in Table~\ref{tbl-fits}.

@@

Figure~\ref{fig-number} shows the SFR as a function of $M_\star$ for disk galaxies separated by their arm multiplicity. The GZ2 data separates disk galaxies with visible spiral arms into categories of 1, 2, 3, 4, or more than four spiral arms; there is also an additional option if the number of spiral arms cannot be accurately determined. For this analysis, galaxies in each mass/SFR bin is weighted by the vote fraction for the morphology being tested. {\rr The fit to the star-forming main sequence for all six sub-samples of spiral galaxies split by arm multiplicity tightly follows that of the total star-forming disk population. Both the slopes and offsets of each linear fit are consistent within the formal fitting errors in Table~\ref{tbl-fits}. %There are too few individual galaxies with confirmed spiral morphologies at lower masses to do a reliable fit; however, the same relation is reproduced for all arm multiplicities at $10^8<M_\star/M_\odot<10^{11}$.

@@

We have repeated the analysis above for the sub-sample of disk galaxies for which the spiral multiplicity is determined with high confidence ($p_\textrm{arms~number} > 0.8$) by GZ2, thus eliminating ``intermediate'' galaxies for which the morphology is uncertain. These galaxies ($N=10,035$) are dominated by two-armed spirals, which are the only spiral multiplicity for which significant numbers of galaxies at $M_\star<10^9~M_\odot$ are detected. The results for the SFMS are qualitatively the same as when using the weighting scheme, although we note that there are too few examples ($N<10$) of either three- or four-armed spirals for a reliable fit. The offset of the one-armed spirals above the SFMS is also significantly more pronounced when using high-confidence morphologies.}

@@

The agreement of all sub-varieties of star-forming galaxies is supported by the close agreement to the linear fits to the data for all well-sampled categories (Table~\ref{tbl-fits}). This tracks only the slope and offset of the distribution, however, and not its width. We thus also compare the sample standard deviation ($\sigma_{SFR}$) to the star-forming galaxy population over its various morphological subsamples. The value of $\sigma_{SFR}$ monotonically decreases with increasing mass over the range {\rr $8.0<\log(M/M_\odot)<11.5$. For all morphological populations examined in this paper}, the widths of their distributions are consistent with the broader population (Figure~\ref{fig-sigma}).

@@

We have {\rr also examined all the populations of galaxies described above (bars, arm pitch angle, arm multiplicity) and measured the differences when using specific star formation rate ($\textrm{sSFR}\equiv\textrm{SFR}/M_\star$) instead of SFR. There is no significant change in any of the morphologically-selected categories as compared to the general SFMS. }

@@

\citet{abr14} found that by normalizing galaxies by the stellar mass of the disk alone, the slope of the SFMS is consistent with only a linear trend (removing any dependence on mass). Although this correction to the disk stellar mass homogenizes the SFMS for disks with a range of $B/T$, the intrinsic dispersion ($\sigma_{SFR}$) of the sequence must be a result of contributions by bars, disk dynamics, halo heating, AGN activity, {\rr environment and/or gas accretion history, among other factors \citep{dut10}. Our results show that the neither of the first two factors play dominant roles in controlling $\sigma_{SFR}$, at least as far as major dynamical drivers (such as strong bars or additional arms) are concerned. Thus while the overall bulge strength does affect the position of a galaxy on the SFMS \citep{mar09,che12a,fan13,kav14,lan14,oma14}, the structure of the \emph{disk itself} does not. This is also consistent with recent models in which details of the feedback, which also relate strongly to the galaxy properties, have little effect on the SFMS \citep{hop14}. Alternatively, this also agrees with models in which the SFMS is the result of stochastic processes, rather than deterministic physics related to galaxy evolution \citep{kel14b}. }

@@

We analyze for the first time the detailed structure of disks in large samples of star-forming galaxies {\rr in the local Universe }as related to their position on the $M_\star$-SFR relation. This analysis is made possible by using morphological classifications from the Galaxy~Zoo~2 project. We find that neither the slope nor the dispersion of the star-forming galaxies are affected when splitting the sample into different categories of disks, including barred/unbarred galaxies, the pitch angle of spiral arms, or the number of spiral arms.
